Insertion Sort Analysis in Java

### 2015-02-04 Programming Project A/Insertion Sort Analysis in Java.tex
finished first draft

%% BEGIN TESTING
\section{Testing}
\subsection{Testing Plan and Results}
All arrays used in testing are Java ArrayList<Integer> unless otherwise specified. All times are recorded in milliseconds using a stopwatch class borrowed from [NEED CITATION]. %% TODO: Add Citation %%
It is important to note that the stopwatch class used takes the elapsed real-time between the start of the insertion sort algorithm and the end of the insertion sort algorithm as opposed to taking the elapsed processor-time because these tests were run on a multi-core computer. 
In the table below, A denotes Array. Times in the table below are given as averages out of 10 trials.
\newline
\captionof{table}{Test Results}
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline Tested Input & Expected Results & Actual Results & Time \\
\hline Empty A & Empty A & Empty A & 0.0003 \\
\hline A of 1000 Strings & Sorted A 1000 Strings & Sorted A 1000 Strings & 0.021 \\
\hline A 1 Element & Original A & Original A & 0.0003 \\
\hline A 10 Elements & Sorted A 10 Elements & Sorted A 10 Elements & 0.0005 \\
\hline A 100 Elements & Sorted A 100 Elements & Sorted A 100 Elements & 0.0021 \\
\hline A 1000 Elements & Sorted A 1000 Elements & Sorted A 1000 Elements & 0.019 \\
\hline A 10000 Elements & Sorted A 10000 Elements & Sorted A 10000 Elements & 0.129 \\
\hline A 100000 Elements & Sorted A 100000 Elements & Sorted A 100000 Elements & 6.4923 \\
\hline A 1000000 Elements & Sorted A 1000000 Elements & Sorted A 1000000 Elements & 2135.5007 \\
\hline A 10000000 Elements & Sorted A 10000000 Elements & OS Crash & N/A \\
\hline A 1000 Identical Elements & Original Array & Original Array & 0.0052 \\
\hline
\end{tabular}
\end{center}

@@

The last issue encountered during the development process of this program was generic programming. The program would not compile because not all Objects have a compareTo method. Adding the statement $<T$ $extends$ $Comparable<T>>$ fixed this issue.

@@

The expected growth of insertion sort as the number of elements (n) grows large can be represented as $ \theta(n^2) $ where $ \theta() $ represents the asymptotically tightly bound running time. At $n_0$ the algorithm took 0.0003 milliseconds to complete. At $n_1$ the algorithm also took 0.0003 milliseconds to complete. For both these values of $n$, the algorithm runs at a constant time because no for-loops are executed. As $n$ grows larger the time to complete the experimental data correlates quite accurately with the expected growth. For $n_{1*10^6} $ the data matches up perfectly. This is to be expected, because as n grows larger the constants and lower orders of the actual running time of the insertion sort start to affect the running time less and less as the highest order of $n^2$ is so large. It is expected for lower values of n to not correlate well with $ \theta(n^2) $ because the system running the algorithm may be executing superfolous commands such as checking for system. For the data graphed above (see Fig. 1) this was in fact the case. To aid this issue, averages of 10 trails were taken and used in Figure 1.
%% END EXPERIMENTAL ANALYSIS

>>>>

%% BEGIN TESTING
\section{Testing}
\subsection{Testing Plan and Results}
All arrays used in testing are Java ArrayList<Integer> unless otherwise specified. All times are recorded in milliseconds using a stopwatch class borrowed from Robert Sedgewick and Kevin Wayne \cite{stopwatch}.
It is important to note that the stopwatch class used takes the elapsed real-time between the start of the insertion sort algorithm and the end of the insertion sort algorithm as opposed to taking the elapsed processor-time because these tests were run on a multi-core computer. 
In the table below, A denotes Array. Times in the table below are given as averages out of 10 trials.
\newline
\captionof{table}{Test Results}
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline Tested Input & Expected Results & Actual Results & Time \\
\hline Empty A & Empty A & Empty A & 0.0003 \\
\hline A of 1000 Strings & Sorted A 1000 Strings & Sorted A 1000 Strings & 0.021 \\
\hline A 1 Element & Original A & Original A & 0.0003 \\
\hline A 10 Elements & Sorted A 10 Elements & Sorted A 10 Elements & 0.0005 \\
\hline A 100 Elements & Sorted A 100 Elements & Sorted A 100 Elements & 0.0021 \\
\hline A 1000 Elements & Sorted A 1000 Elements & Sorted A 1000 Elements & 0.019 \\
\hline A 10000 Elements & Sorted A 10000 Elements & Sorted A 10000 Elements & 0.129 \\
\hline A 100000 Elements & Sorted A 100000 Elements & Sorted A 100000 Elements & 6.4923 \\
\hline A 1000000 Elements & Sorted A 1000000 Elements & Sorted A 1000000 Elements & 2135.5007 \\
\hline A 10000000 Elements & Sorted A 10000000 Elements & OS Crash & N/A \\
\hline A 1000 Identical Elements & Original Array & Original Array & 0.0052 \\
\hline
\end{tabular}
\end{center}

@@

The last issue encountered during the development process of this program was generic programming. The program would not compile because not all Objects have a compareTo method. Adding the statement $<T$ $extends$ $Comparable<T>>$ fixed this issue. \cite{erik}

@@

The expected growth of insertion sort as the number of elements (n) grows large can be represented as $ \theta(n^2) $ where $ \theta() $ represents the asymptotically tightly bound running time. \cite{textbook} At $n_0$ the algorithm took 0.0003 milliseconds to complete. At $n_1$ the algorithm also took 0.0003 milliseconds to complete. For both these values of $n$, the algorithm runs at a constant time because no for-loops are executed. As $n$ grows larger the time to complete the experimental data correlates quite accurately with the expected growth. For $n_{1*10^6} $ the data matches up perfectly. This is to be expected, because as n grows larger the constants and lower orders of the actual running time of the insertion sort start to affect the running time less and less as the highest order of $n^2$ is so large. It is expected for lower values of n to not correlate well with $ \theta(n^2) $ because the system running the algorithm may be executing superfolous commands such as checking for system. For the data graphed above (see Fig. 1) this was in fact the case. To aid this issue, averages of 10 trails were taken and used in Figure 1.
%% END EXPERIMENTAL ANALYSIS
